Types From Data: \textnormal{Making structured schema-free data formats first-class citizens in F\#}

### 2015-11-17 pldi-2016/paper.tex
Update paper.tex

\item The inference mechanism prefers types with properties (records). This is to allow extensible (open-world)
  data formats, as we will explain. It also interacts well with developer tooling --
  most editors provide code completion hints on ``.'' and so types with properties (records)
  are easier to use than typessuch as discriminated unions that require pattern matching.

@@

\paragraph{The role of records.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
are available for a person. To support this scenario, our inference algorithm prefers records 
(we also treat XML elements and CSV rows as records) and only infers types that require pattern 
matching as the last resort.

In the above example, this is demonstrated by the fact that \ident{Age} is marked as optional.
An alternative is to provide two different record types (one with \ident{Name} and other with 
\ident{Name} and \ident{Age}), but this would complicate the processing code.

@@

In the above example, we do not infer an F\# discriminated union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. 
The type is not closed -- the data might be some other element about which nothing is known.
This is the \emph{open-world assumption}. Supporting the open-world assumption is crucial when reading data from 
external data sources, because the sample document need only
mention possibilities relevant to the application.

@@

In this case, the variant-annotated top type is ultimamtely presented as the following F\# type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
This still provides access to the possible elements known statically from the sample.
On the other hand, the user also needs to explicitly handle the casse when the value is none 
of the statically known elements. 
For example, the above code for  \strf{sample.doc} includes a \kvd{None -> ()} case to ensure that it will work 
correctly on a document containing elements not listed in the sample. Similarly, 
the program is not \emph{forced} to consider each of the possiblities in the annotations (in the manner
of complete-coverage checks for discriminated unions). For example,
the program is not forced to cover cases for  \ident{Paragraph} and \ident{Image}.  This is consistent
with our design principles, which prefer records (with optional properties) over discriminated unions.

@@

A variant type \kvd{any} is annotated with the types it may represent in the sample document.
As discussed earlier (\S\ref{sec:providers-xml}), variants force the user to handle the case when 
a value is of a different type than the statically known ones and so variant types can be seen as a family of 
top types. For the same reason, variants also implicitly permit the \kvd{null} value.

@@

Given two structural types, the \emph{common supertype} relation finds a least upper bound
(Theorem~\ref{thm:lub}). When possible, it avoids variant types and prefers records 
(Corollary~\ref{thm:no-unions}), which is important for usability as discussed earlier
(\S\ref{sec:providers-json}).

@@

As discussed earlier, variant types are also generated as F\# classes with properties. Given a variant
$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
a runtime type test using \ident{hasType} to ensure that the value has the right type (similarly to runtime 
type conversions from the top type in languages like Java). If the type matches, a \ident{Some} value is 
returned. The type inference algorithm also guarantees that there is only one case for each type tag 
(\S\ref{sec:inference-commonsuper}) and so we can use the tag when naming the generated member 
(using the \ident{nameof} function).

>>>>

\item The inference mechanism prefers class types with properties. This is to allow extensible (open-world)
  data formats, as we will explain. It also interacts well with developer tooling --
  most editors provide code completion hints on ``.'' and so class types with properties
  are easier to use than types such as discriminated unions that require pattern matching.

@@

\paragraph{The role of optional properties.}
The sample code is easy to write thanks to the fact that most F\# editors provide code completion
when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
are available for a person. To support this scenario, our inference algorithm prefers types with properties
and only infers discriminated union types that require pattern matching as the last resort.

In the above example, this is demonstrated by the fact that \ident{Age} is marked as optional.
An alternative is to provide a discriminated union for two different record types (one with \ident{Name} and other with 
\ident{Name} and \ident{Age}), but this would complicate the processing code.

@@

In the above example, we do not infer an F\# discriminated union type representing a choice between 
heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
contain elements prefixed with other namespaces.}. 
The type is not closed -- the data might be some other element about which nothing is known.
This is the \emph{open-world assumption}. Supporting the open-world assumption is useful when reading data from 
external data sources, because the sample document need only
mention possibilities relevant to the application.

@@

In this case, the possibilities are ultimately presented as the following F\# class type:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Element}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
 \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
\end{array}
\end{equation*}
%
This still provides access to the possible elements known statically from the sample.
However this  access is deliberately `weak', because the user needs to explicitly handle 
the case when a value is not of a statically known element. 
For example, the above code for  \strf{sample.doc} includes a \kvd{None -> ()} case to ensure that it will work 
correctly on a document containing elements not listed in the sample. Furthermore, 
the programmer is not forced to consider each of the possiblities in the annotations
such as \ident{Paragraph} and \ident{Image}.  This is  consistent with our design principles, 
which prefer optional properties over discriminated unions. The gain is that that the provided types support 
both open-world data and dot-driven developer tooling.

@@

A variant type \kvd{any} is annotated with the types it may represent in the sample document.
Variant-annotated top types can be seen as a family of top types and implicitly permit the \kvd{null} value.

@@

Given two structural types, the \emph{common supertype} relation finds a least upper bound
(Theorem~\ref{thm:lub}). When possible, it avoids variant-annotated top types and prefers records 
(Corollary~\ref{thm:no-unions}), which is important for usability as discussed earlier
(\S\ref{sec:providers-json}).

@@

As discussed earlier, variant-annotated top types are generated as F\# classes with properties. Given a variant
$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
a runtime type test using \ident{hasType} to ensure that the value has the right type (similarly to runtime 
type conversions from the top type in languages like Java). If the type matches, a \ident{Some} value is 
returned. The type inference algorithm also guarantees that there is only one case for each type tag 
(\S\ref{sec:inference-commonsuper}) and so we can use the tag when naming the generated member 
(using the \ident{nameof} function).
